Do moral communities have a spatial dimension? A spatial exploratory analysis of places of worship and violent crime in the city of Recife, Brazil

### 2019-10-16 Moral_Communities_and_Crime_v1.Rmd
Initial submission to Applied Spatial Analysis and Policy

<!-- This paper is an R markdown document that includes self-contained, reproducible analysis. To reproduce the paper the following file is needed: `Moral_Communities_and_Crime.RData`. For reference: this data file was created using R notebook `00. Data Preprocessing.Rmd`. The initial analysis was done in R notebook `01. Analysis.Rmd`. Neither of these two files are needed for this document. -->

@@

This concern is not iddle: the endemic malady of violent crime represents an important factor that threatens to derail sustainable development goals. Violent crime, furthermore, tends to be more accute in those regions most in need of development. On the one hand, violent crime represents an economic and social drag [@Becker2017analise]. The impacts are multifaceted, since they affect people's well-being, either through the loss of human life, mental health, and limitations in the right to public spaces [@Doran2011putting], or through disturbances in schooling and academic achievement [@Unodc2019development]. In turn, these negative effects combine to create the unfortunate conditions that tend to breed criminal behavior, thus creating a vicious cycle of economic disadvantage and violence [@Unodc2019development]. It is not surprising, given these pecularities, that researchers have joined calls for increased attention to the study of the patterns of violent crime in Low and Middle Income Countries [LMIC; @Murray2013crime].

Added to the scenario, formal social control institutions in countries like Brazil leave something (or much) to be desired: while in practice they should inhibit criminal behavior, they suffer from deep deficiencies that end up reducing the deterrent power of the justice system. Serious institutional problems include the inefficiency of the police, the lack of national legislation, the glacial pace of judicial processes, and the weak situation of the prison system in the country [@Menezes2013spatial]. This makes it even more urgent to understand the role of _informal social controls_, i.e., the ability of community organizations such as schools, clubs, and neighborbood associations to suppress crime by strengthening the capabilities of neighbors to control inappropriate behavior [@Groff2015informal].

In order to understand the factors that can potentially help to deter crime, beyond formal control, it is important to identify empirical regularities. Criminological factors include concrete elements (such as te presence of arms or drugs, or elements of the built environment), and also figurative factors, which include the soci costs of deviant behavior as well as family supervision. Accordingly, a number of studies have investigated various aspects of the environment and neighborhood design [e.g., @Foster2010neighbourhood; @He2017built; @Loukaitou2001measuring], whereas other studies have focused on exposure to environmental attributes that signal weakened norms, such as liquor and tobacco outlets [e.g., @Brower2007spatial; @Deryol2016crime; @Lipton2008spatial; @Quick2017influence].

@@

With the above considerations in mind, the objective of this paper is to investigate whether and how places of workship correlate spatially with criminal events, or in other words, to investigate whether their signals as moral communities have a discernible spatial dimension. The case study is the city of Recife in the state of Pernambuco, in Brazil's Northeast. Recife is a large and important metropolitan area in a historically poor region, and afflicted by high levels of violent crime, having the dubious honor of being one of five state capitals with the highest rates of homicide in the period under study [@Menezes2013spatial]. The empirical strategy is to use spatial analysis to explore the potential geographical relationships between violent criminal events, on the one hand, and places of worship and a selection of commercial establishments that serve as controls, on the other. Dissagregated data allows us to analyze the relevant point patterns at a very high level of resolution.

@@

An early effort to theorize the effect of religion on criminal behavior was the _hellfire hypothesis_ of Hirschi and Stark [-@Hirschi1969hellfire]. The focus of this hypothesis is the threat of extratemporal (and possibly eternal) punishment, and how this threat can deter believers from commandment-breaking (also see Pascal's wager). Accordingly, Hirsch and Stark [-@Hirschi1969hellfire] posit that negative correlations between crime and religion are a consequence of a sense of commitment with normative values, a commitment that is ritually reinforced in a regular fashion (e.g., the liturgical rite of peace in Catholic Mass). Rohrbaugh and Jessor [-@Rohrbaugh1975religiosity], for instance, note that by attributing to divinity (or some other supra-terrenal entity) the supreme force of punishment, religion helps to build a sense of obduracy, or "hardening" against temptation.

@@

The hypothesis of moral communities has over the years been used to examine a variety of outcomes of interest. Recent examples include Stroope and Baker's [-@Stroope2018moral] exploration of religiosity and self-rated health; Davignon and Thomson [-@Davignon2015christian] with their research on institutional context and the religiosity of students; and the study of religion as a source of trust of Traunmuller [-@Traunmuller2011moral]. This is in addition, of course, to numerous studies on criminal behavior such as Eitle [-@Eitle2011religion], an author who explored the deterrence power of religion on gambling; Lee and Bartkowski's [-@Lee2004love] investigation of juvenile homicide in rural areas; and the research of Regnerus [-@Regnerus2003moral] on adolescent delinquency. The hypothesis of moral communities as a form of informal social control, on the other hand, has been less studied from a geographical perspective, and it is only recently that has attracted the attention of researchers. Groff [-@Groff2015informal], for instance, discusses informal social control as a phenomenon that can plaussibly operate at different geographical scales, from the level of the home and family, through the street block and neighborhood, possibly to other scales such as the county. In this way, Nie and Yang [-@Nie2019smoking] remark on a recent paper on the lack of research conducted to study how the religious context of a geographical area (e.g., a county) may influence (youth smoking) behavior (p. 2). This point is echoed by Warner and Konkel [-@Warner2019neighborhood], who moreover recommend the use of even smaller units of analysis (e.g., Census Block Groups), in an effort to reduce aggregation bias [see @Hipp2007block], but more importantly because the processes defined by social disorganization theory are thought to occur at relatively small geographical scales.

@@

Using quadrats of equal size is a convenient simplication, but in principle the areas could be different - in which case the count of events would be proportional to the area of the quadrat under homogeneity. An interesting variation of this technique is conditional quadrat counts, whereby the partition of regions $\mathcal{A}_1,\cdots,\mathcal{A}_m$ is done to reflect an underlying covariate of interest, say $Z$. By introducing a covariate as a partitioning criterion, it becomes possible to calculate estimators of the intensity of quadrats at different levels of the value of the covariate. For exploratory purposes, we can plot average intensity by quadrat, and compare to the global intensity.

@@

```{r table-descriptive-statistics, echo=FALSE}
descriptive_stats <- data.frame(Attribute = c("Man",
                                              "Woman",
                                              "Murder",
                                              "Robbery",
                                              "Body injury followed by death",
                                              "Black and White",
                                              "Yellow and White",
                                              "Not Reported",
                                              "1 to 12 years",
                                              "13 to 17 years",
                                              "18 to 30 years",
                                              "31 to 65 years",
                                              "65 years and older",
                                              "Not Reported",
                                              "Firearm",
                                              "Other"),
                                Percentage = c(91.35,
                                               8.65,
                                               97.82,
                                               2.00,
                                               0.18,
                                               95.11,
                                               1.27,
                                               3.62,
                                               0.18,
                                               11.31,
                                               61.70,
                                               25.84,
                                               0.73,
                                               0.24,
                                               87.51,
                                               12.49))

@@

In addition to places of worship, the National Register of Addresses for Statistical Purposes was queried to extract facilities other than places of worship. As discussed above, the idea is to identify points of reference that can be used as controls, having a neutral morality profile. For the sake of the present study, we selected pharmacies, ice cream shops, and bakeries to construct our control group. These three types of establishments comply with the criteria of being morally neutral and having a spatial distribution commensurate with places of worship.
Figure 2 shows the spatial distribution of the control establishments, namely pharmacies, bakeries, and ice cream shops in the city of Recife. Note the similarity between the maps. As expected, there are differences in the number of points, but the locations of cases and controls are quite similar.

@@

Figure \ref{fig:plot-sed-as-quintiles} displays the geography of Socio-Economic Deprivation in Recife, after classifying _setores_ by quintiles, whereby the "High" class corresponds to _setores_ in the top 20% of the Socio-Economic Deprivation indicator, and the "Low" class corresponds to _setores_ in the bottom 20% of the Socio-Economic Deprivation indicator. The figure shows a veritable mosaic of affluence and deprivation, with high deprivation areas directly in contact, and in some cases almost completely surrounded, by low deprivation areas. This geographical pattern of unequality, on the other hand, seems to be characteristic of Brazilian metropolitan regions, where enclaves of wealth and _favelas_ (i.e., urban slums) can be found in close proximity [see for example @Feitosa2007global].

@@

Figure \ref{fig:plot-crime-quadrat} shows the results of conditional quadrat counts for Lethal and Intentional Violent Crime. The dotted line indicates the global intensity, simply the total number of criminal events divided by the area of the region. It can be seen that the intensity of crime clearly increases with Socio-Economic Deprivation, and the intensity in areas with the highest levels of Socio-Economic Deprivation is `r crime5_intensity[5]/crime5_intensity[1] * 100` % higher than in places with the lowest levels of Socio-Economic Deprivation. This is as expected, as research has consistently found correlations between Socio-Economic Deprivation and a variety of negative health outcomes, including homicide [@Ichihara2018area]. A $\chi^2$ test of independence on the quadrat counts yields a value of $p\leq 0.0001$ (four degrees of freedom), which comfortably rejects the null hypothesis of homogeneity of intensity of crime by level of Socio-Economic Deprivation.

@@

Three types of commercial establishments also appear to display inhomogeneous locational patterns (Table \ref{tab:quadrat-test-statistics}), with the $\chi^2$ test of independence emphaticaly rejecting the null hypothesis. As seen in Figure \ref{fig:plot-events-quadrat}, ice cream shops tend to be found more frequently in places with middle-low Socio-Economic Deprivation, and less frequently in places with middle-high Socio-Economic Deprivation. Pharmacies show a clearer trend, with locational patterns that tend to favor places with low Socio-Economic Deprivation. Finally, supermarkets are found less frequently in places with middle-high and high Socio-Economic Deprivation.

@@

quad.tests <- data.frame(Event = c("Catholic",
                                   "Evangelical",
                                   "Spiritist",
                                   "Ice Crem Shop",
                                   "Pharmacy",
                                   "Supermarket"),
                         rbind(quadrat.test(catholic5), 
                               quadrat.test(evangelic5), 
                               quadrat.test(spiritist5),
                               quadrat.test(ice_cream5),
                               quadrat.test(pharmacies5),
                               quadrat.test(super_markets5))) %>%
  transmute(Event,
            Statistic = as.numeric(statistic),
            degrees_of_freedom = as.numeric(parameter),
            p_value = as.numeric(p.value)) %>%
  mutate(p_value = ifelse(p_value >= 0.001, paste0(round(p_value, 4)), "$<0.001$"))

@@

The results of the analysis are shown in Figure \ref{fig:plot-relative-distribution-with-baseline}. The general trends shown by the relative distribution functions are similar to those seen above, before the introduction of a baseline function. We notice that at short distances the intensity of crime as a function of distance to the various places of worship and commercial establishments is higher than the baseline intesity (i.e., the intensity according to Socio-Ecomomic Deprivation). The only exception is intensity as a function of distance to a Spiritist church, which at short distances is not significantly different from the baseline intensity (see the confidence bands).

The peak intensity as a function of distance to Evangelical churches is `r round(max_int_evangelical_bl, 2)` times the baseline intensity; for Catholic places of worship, the highest intesity is `r round(max_int_catholic_bl, 2)` times the baseline intensity; and the intensity reaches a peak of `r round(max_int_spiritist_bl, 2)` times the baseline intensity when the covariate is distance to Spiritist churches. Turning now to inspect the three types of commercial establishments, we see that the highest intensities are `r round(max_int_ice_cream_bl, 2)`, `r round(max_int_pharmacy_bl, 2)`, and `r round(max_int_supermarket_bl, 2)` times the intensity according to Socio-Economic Deprivation for ice cream shops, pharmacies, and supermarkets, respectively.

@@

In this respect, it has been argued that Catholics and some Protestant groups are more adept at generating bridging social capital with stron intergroup relations. This can result in greater collective efficacy - which in turn can decrease the likelihood of crime [see @Hoffmann2015religion, p. 2]. Along these lines, Beyerlein and Hipp [-@Beyerlein2005social], examined data for 3,157 counties in the United States in 2000, and found that counties with greater percentages of people affiliated with mainline Protestant and Catholic traditions tended to have lower crime rates. Similarly, Warner and Konkel [-@Warner2019neighborhood] found that mainline Protestant and briding churches (but not Catholic churches) correlated positively with informal social control at the neighborhood level in two cities in Kentucky. Nie and Yang [-@Nie2019smoking], in a study of smoking, also found that share of Catholic population in counties did not reduce the rate of smoking, thus bringing into question the efficacy of this type of moral community in generating social control - albeit at a fairly aggregated scale (counties). Our findings with respect to Catholic places of worship echo these results at a much higher level of disaggregation, as we fail to find evidence that Catholic churches project a moral community in space more than any of various common types of business establishments.

Conservative Protestant and Evangelic traditions, unlike mainline Protestants and Catholics, appear to focus less on bridging and more on bonding social capital with strong within-group relations. Associated with this, Conservative Protestant doctrine emphasizes individual responsibility, and therefore social ills tend to be seen as primarily personal ills in need of religious redemption instead of secular interventions [@Nie2019smoking, pp. 2-3]. Under this light, a moral community with such views of social problems may not be effective to deter crime [@Ellison2003enduring]; instead, conservative Protestant doctrine may be more tolerant of violent behavior when associated with defense of honor, family, property, or women, or unfortunate events may be seen as representing legitimate celestial retribution for moral torpitude. Empirical research by Beyerlein and Hipp [-@Beyerlein2005social] found that counties with greater percentages of people affiliated with Evangelical traditions tended to have higher crime rates. More recent research by Desmond et al. [-@Desmond2010congregations] in more than 400 block groups in Indianapolis found that neighborhoods with a greater presence of Evangelical congregations had higher rates of both violent and property crimes. In the specific context of Brazil, it is important to consider the role that religion has had in interpreting and conferring meaning to urban violence, or what Brazilian anthropologist Patricia Birman has termed "The Violence of the Just" [see @Birman2012violencia; @Birman2019narrativas].

Our analysis suggests that the role of Evangelical churches may be more complex than thought: on the one hand, our results indicate that Lethal and Intentional Violent Crime tends to be more intense in the proximity of Evangelical places of worship, up to a distance of approximately 100 m, but then declines even below the global intensity and the background intenstiy at distances between 300 m and 100 m. It is at the moment unclear what could cause this, and candidate explanations could include the way police respond to crime incidents in different neighborhoods [e.g., Warner, 1997], the likelihood of residents reporting crime to the police (e.g., Goudriaan, Witterbrood, & Niewbeerta, 2006), and the level of opportunity for crimes (e.g., Rountree, Land, & Miethe, 1994). Could Evangelicals churches provide greater levels of opportunity for crime, if adherents to this tradition are seen as meek? And is it possible that at a different scale, unlike reports by other researchers, Evangelical churches are more effective at keeping social control? While these speculations cannot be answered in the context of the present study, we suggest that these might be fruitful areas for future research.

Finally, we would like to remark on the results regarding Spiritist churches, which in some way are the mirror image of those for Evangelical churches. An extensive scan of the literature finds much less information about this religious tradition, in particular in the context of Brazil where it adapted in ways that made it very different from its original French predecessor [see @daGraca2011espiritismo]. Could it be that Spiritism is seen by potential criminals as spiritually or materially risky, if supernatural retribution is feared? Again, we cannot do more than speculate about this at the moment, but this might be another worthwhle avenue for future investigation.

@@

1. Catholic places of worhsip do not seem to geographically project more or less of a moral community effect than, say, ice cream shops, pharmacies, or supermarkets: the intensity of crime as a function of distance to Catholic churches is not significantly different that what we observe when we consider distance to any of these types of commercial establishments, at any distance.

2. The intensity of crime with respect to distance to Evangelical churches is more complex than previously known: whereas the intensity of crime is higher at relativey short distance from Evangelical places of worship, at longer distances it tends to decline even below the global intensity of crime and the background level of crime according to Socio-Economic Deprivation.

@@

In summary, the research presented on this paper provides information about the potential of different types of moral communities to reduce crime, and should be of interest to policy makers as they assess whether formal or informal forms of social control can be effective to deter criminal behavior.

>>>>

<!-- This paper is an R markdown document that includes self-contained, reproducible analysis. To reproduce the paper the following files are needed:

1. Moral_Communities_and_Crime_v1.Rmd (this R markdown document)

2. `Moral_Communities_and_Crime_Recife.RData`. For reference: this data file was created using R notebook `00. Data Preprocessing.Rmd`. The initial analysis was done in R notebook `01. Analysis.Rmd`. Neither of these two files are needed for this document.

3. For compiling into a pdf in the style of a Springer document: svjour3.cls, svglov3.clo, spphys.bst, spmpsci.bst, spbasic.bst

-->

@@

This concern is not idle: the endemic malady of violent crime represents an important factor that threatens to derail sustainable development goals. Violent crime, furthermore, tends to be more acute in those regions most in need of development. On the one hand, violent crime represents an economic and social drag [@Becker2017analise]. The impacts are multifaceted, since they affect people's well-being, either through the loss of human life, mental health, and limitations in the right to public spaces [@Doran2011putting], or through disturbances in schooling and academic achievement [@Unodc2019development]. In turn, these negative effects combine to create the unfortunate conditions that tend to breed criminal behavior, thus creating a vicious cycle of economic disadvantage and violence [@Unodc2019development]. It is not surprising, given these peculiarities, that researchers have joined calls for increased attention to the study of the patterns of violent crime in Low and Middle Income Countries [LMIC; @Murray2013crime].

Added to the scenario, formal social control institutions in countries like Brazil leave something (or much) to be desired: while in practice they should inhibit criminal behavior, they suffer from deep deficiencies that end up reducing the deterrent power of the justice system. Serious institutional problems include the inefficiency of the police, the lack of national legislation, the glacial pace of judicial processes, and the weak situation of the prison system in the country [@Menezes2013spatial]. This makes it even more urgent to understand the role of _informal social controls_, i.e., the ability of community organizations such as schools, clubs, and neighborhood associations to suppress crime by strengthening the capabilities of neighbors to control inappropriate behavior [@Groff2015informal].

In order to understand the factors that can potentially help to deter crime, beyond formal control, it is important to identify empirical regularities. Criminological factors include concrete elements (such as the presence of arms or drugs, or elements of the built environment), and also figurative factors, which include the social cost of deviant behavior. Accordingly, a number of studies have investigated various aspects of the environment and neighborhood design [e.g., @Foster2010neighbourhood; @He2017built; @Loukaitou2001measuring], whereas other studies have focused on exposure to environmental attributes that signal weakened norms, such as liquor and tobacco outlets [e.g., @Brower2007spatial; @Deryol2016crime; @Lipton2008spatial; @Quick2017influence].

@@

With the above considerations in mind, the objective of this paper is to investigate whether and how places of worship correlate spatially with criminal events, or in other words, to investigate whether their signals as moral communities have a discernible spatial dimension. The case study is the city of Recife in the state of Pernambuco, in Brazil's Northeast. Recife is a large and important metropolitan area in a historically poor region, and afflicted by high levels of violent crime, having the dubious honor of being one of five state capitals with the highest rates of homicide in the period under study [@Menezes2013spatial]. The empirical strategy is to use spatial analysis to explore the potential geographical relationships between violent criminal events, on the one hand, and places of worship and a selection of commercial establishments that serve as controls, on the other. Dissagregated data allows us to analyze the relevant point patterns at a very high level of resolution.

@@

An early effort to theorize the effect of religion on criminal behavior was the _hellfire hypothesis_ of Hirschi and Stark [-@Hirschi1969hellfire]. The focus of this hypothesis is the threat of extra-temporal (and possibly eternal) punishment, and how this threat can deter believers from commandment-breaking (also see Pascal's wager). Accordingly, Hirsch and Stark [-@Hirschi1969hellfire] posit that negative correlations between crime and religion are a consequence of a sense of commitment with normative values, a commitment that is ritually reinforced in a regular fashion (e.g., the liturgical rite of peace in Catholic Mass). Rohrbaugh and Jessor [-@Rohrbaugh1975religiosity], for instance, note that by attributing to divinity (or some other supra-terrenal entity) the supreme force of punishment, religion helps to build a sense of obduracy, or "hardening" against temptation.

@@

The hypothesis of moral communities has over the years been used to examine a variety of outcomes of interest. Recent examples include Stroope and Baker's [-@Stroope2018moral] exploration of religiosity and self-rated health; Davignon and Thomson [-@Davignon2015christian] with their research on institutional context and the religiosity of students; and the study of religion as a source of trust of Traunmuller [-@Traunmuller2011moral]. This is in addition, of course, to numerous studies on criminal behavior such as Eitle [-@Eitle2011religion], an author who explored the deterrence power of religion on gambling; Lee and Bartkowski's [-@Lee2004love] investigation of juvenile homicide in rural areas; and the research of Regnerus [-@Regnerus2003moral] on adolescent delinquency. The hypothesis of moral communities as a form of informal social control, on the other hand, has been less studied from a geographical perspective, and it is only recently that has attracted the attention of researchers. Groff [-@Groff2015informal], for instance, discusses informal social control as a phenomenon that can plausibly operate at different geographical scales, from the level of the home and family, through the street block and neighborhood, possibly to other scales such as the county. In this way, Nie and Yang [-@Nie2019smoking] remark on a recent paper on the lack of research conducted to study how the religious context of a geographical area (e.g., a county) may influence (youth smoking) behavior (p. 2). This point is echoed by Warner and Konkel [-@Warner2019neighborhood], who moreover recommend the use of even smaller units of analysis (e.g., Census Block Groups), in an effort to reduce aggregation bias [see @Hipp2007block], but more importantly because the processes defined by social disorganization theory are thought to occur at relatively small geographical scales.

@@

Using quadrats of equal size is a convenient simplification, but in principle the areas could be different - in which case the count of events would be proportional to the area of the quadrat under homogeneity. An interesting variation of this technique is conditional quadrat counts, whereby the partition of regions $\mathcal{A}_1,\cdots,\mathcal{A}_m$ is done to reflect an underlying covariate of interest, say $Z$. By introducing a covariate as a partitioning criterion, it becomes possible to calculate estimators of the intensity of quadrats at different levels of the value of the covariate. For exploratory purposes, we can plot average intensity by quadrat, and compare to the global intensity.

@@

```{r table-descriptive-statistics, echo=FALSE}
descriptive_stats <- data.frame(Attribute = c("Man",
                                              "Woman",
                                              "Murder",
                                              "Robbery",
                                              "Body injury followed by death",
                                              "Black and Brown",
                                              "Yellow and White",
                                              "Not Reported",
                                              "1 to 12 years",
                                              "13 to 17 years",
                                              "18 to 30 years",
                                              "31 to 65 years",
                                              "65 years and older",
                                              "Not Reported",
                                              "Firearm",
                                              "Other"),
                                Percentage = c(91.35,
                                               8.65,
                                               97.82,
                                               2.00,
                                               0.18,
                                               95.11,
                                               1.27,
                                               3.62,
                                               0.18,
                                               11.31,
                                               61.70,
                                               25.84,
                                               0.73,
                                               0.24,
                                               87.51,
                                               12.49))

@@

In addition to places of worship, the National Register of Addresses for Statistical Purposes was queried to extract facilities other than places of worship. As discussed above, the idea is to identify points of reference that can be used as controls, having a neutral morality profile. For the sake of the present study, we selected pharmacies, ice cream shops, and bakeries to construct our control group. These three types of establishments comply with the criteria of being morally neutral and having a spatial distribution commensurate with places of worship.

Figure \ref{fig:plot-events} shows the spatial distribution of the places of worhsip and control establishments in the city of Recife. Note the similarity between the maps. As expected, there are differences in the number of points, but the locations of cases and controls are quite similar.

@@

Figure \ref{fig:plot-sed-as-quintiles} displays the geography of Socio-Economic Deprivation in Recife, after classifying _setores_ by quintiles, whereby the "High" class corresponds to _setores_ in the top 20% of the Socio-Economic Deprivation indicator, and the "Low" class corresponds to _setores_ in the bottom 20% of the Socio-Economic Deprivation indicator. The figure shows a veritable mosaic of affluence and deprivation, with high deprivation areas directly in contact, and in some cases almost completely surrounded, by low deprivation areas. This geographical pattern of inequality, on the other hand, seems to be characteristic of Brazilian metropolitan regions, where enclaves of wealth and _favelas_ (i.e., urban slums) can be found in close proximity [see for example @Feitosa2007global].

@@

Figure \ref{fig:plot-crime-quadrat} shows the results of conditional quadrat counts for Lethal and Intentional Violent Crime. The dotted line indicates the global intensity, simply the total number of criminal events divided by the area of the region. It can be seen that the intensity of crime clearly increases with Socio-Economic Deprivation, and the intensity in areas with the highest levels of Socio-Economic Deprivation is `r round(crime5_intensity[5]/crime5_intensity[1] * 100, 2)`% higher than in places with the lowest levels of Socio-Economic Deprivation. This is as expected, as research has consistently found correlations between Socio-Economic Deprivation and a variety of negative health outcomes, including homicide [@Ichihara2018area]. A $\chi^2$ test of independence on the quadrat counts yields a value of $p\leq 0.0001$ (four degrees of freedom), which comfortably rejects the null hypothesis of homogeneity of intensity of crime by level of Socio-Economic Deprivation.

@@

Three types of commercial establishments also appear to display inhomogeneous locational patterns (Table \ref{tab:quadrat-test-statistics}), with the $\chi^2$ test of independence emphatically rejecting the null hypothesis. As seen in Figure \ref{fig:plot-events-quadrat}, ice cream shops tend to be found more frequently in places with middle-low Socio-Economic Deprivation, and less frequently in places with middle-high Socio-Economic Deprivation. Pharmacies show a clearer trend, with locational patterns that tend to favor places with low Socio-Economic Deprivation. Finally, supermarkets are found less frequently in places with middle-high and high Socio-Economic Deprivation.

@@

quad.tests <- data.frame(Event = c("Catholic",
                                   "Evangelical",
                                   "Spiritist",
                                   "Ice Cream Shop",
                                   "Pharmacy",
                                   "Supermarket"),
                         rbind(quadrat.test(catholic5), 
                               quadrat.test(evangelic5), 
                               quadrat.test(spiritist5),
                               quadrat.test(ice_cream5),
                               quadrat.test(pharmacies5),
                               quadrat.test(super_markets5))) %>%
  transmute(Event,
            Statistic = as.numeric(statistic),
            degrees_of_freedom = as.numeric(parameter),
            p_value = as.numeric(p.value)) %>%
  mutate(p_value = ifelse(p_value >= 0.001, paste0(round(p_value, 4)), "$<0.001$"))

@@

The results of the analysis are shown in Figure \ref{fig:plot-relative-distribution-with-baseline}. The general trends shown by the relative distribution functions are similar to those seen above, before the introduction of a baseline function. We notice that at short distances the intensity of crime as a function of distance to the various places of worship and commercial establishments is higher than the baseline intensity (i.e., the intensity according to Socio-Economic Deprivation). The only exception is intensity as a function of distance to a Spiritist church, which at short distances is not significantly different from the baseline intensity (see the confidence bands).

The peak intensity as a function of distance to Evangelical churches is `r round(max_int_evangelical_bl, 2)` times the baseline intensity; for Catholic places of worship, the highest intensity is `r round(max_int_catholic_bl, 2)` times the baseline intensity; and the intensity reaches a peak of `r round(max_int_spiritist_bl, 2)` times the baseline intensity when the covariate is distance to Spiritist churches. Turning now to inspect the three types of commercial establishments, we see that the highest intensities are `r round(max_int_ice_cream_bl, 2)`, `r round(max_int_pharmacy_bl, 2)`, and `r round(max_int_supermarket_bl, 2)` times the intensity according to Socio-Economic Deprivation for ice cream shops, pharmacies, and supermarkets, respectively.

@@

In this respect, it has been argued that Catholics and some Protestant groups are more adept at generating bridging social capital with strong intergroup relations. This can result in greater collective efficacy - which in turn can decrease the likelihood of crime [see @Hoffmann2015religion, p. 2]. Along these lines, Beyerlein and Hipp [-@Beyerlein2005social], examined data for 3,157 counties in the United States in 2000, and found that counties with greater percentages of people affiliated with mainline Protestant and Catholic traditions tended to have lower crime rates. Similarly, Warner and Konkel [-@Warner2019neighborhood] found that mainline Protestant and bridging churches (but not Catholic churches) correlated positively with informal social control at the neighborhood level in two cities in Kentucky. Nie and Yang [-@Nie2019smoking], in a study of smoking, also found that share of Catholic population in counties did not reduce the rate of smoking, thus bringing into question the efficacy of this type of moral community in generating social control - albeit at a fairly aggregated scale (counties). Our findings with respect to Catholic places of worship echo these results at a much higher level of disaggregation, as we fail to find evidence that Catholic churches project a moral community in space more than any of various common types of business establishments.

Conservative Protestant and Evangelic traditions, unlike mainline Protestants and Catholics, appear to focus less on bridging and more on bonding social capital with strong within-group relations. Associated with this, Conservative Protestant doctrine emphasizes individual responsibility, and therefore social ills tend to be seen as primarily personal ills in need of religious redemption instead of secular interventions [@Nie2019smoking, pp. 2-3]. Under this light, a moral community with such views of social problems may not be effective to deter crime [@Ellison2003enduring]; instead, conservative Protestant doctrine may be more tolerant of violent behavior when associated with defense of honor, family, property, or women, or unfortunate events may be seen as representing legitimate celestial retribution for moral turpitude. Empirical research by Beyerlein and Hipp [-@Beyerlein2005social] found that counties with greater percentages of people affiliated with Evangelical traditions tended to have higher crime rates. More recent research by Desmond et al. [-@Desmond2010congregations] in more than 400 block groups in Indianapolis found that neighborhoods with a greater presence of Evangelical congregations had higher rates of both violent and property crimes. In the specific context of Brazil, it is important to consider the role that religion has had in interpreting and conferring meaning to urban violence, or what Brazilian anthropologist Patricia Birman has termed "The Violence of the Just" [see @Birman2012violencia; @Birman2019narrativas].

Our analysis suggests that the role of Evangelical churches may be more complex than thought: on the one hand, our results indicate that Lethal and Intentional Violent Crime tends to be more intense in the proximity of Evangelical places of worship, up to a distance of approximately 100 m, but then declines even below the global intensity and the background intensity at distances between 300 m and 100 m. It is at the moment unclear what could cause this, and candidate explanations could include the way police respond to crime incidents in different neighborhoods, the likelihood of residents reporting crime to the police, and the level of opportunity for crimes [@Warner2019neighborhood]. Could Evangelicals churches provide greater levels of opportunity for crime, if adherents to this tradition are seen as meek? And is it possible that at a different scale, unlike reports by other researchers, Evangelical churches are more effective at keeping social control? While these speculations cannot be answered in the context of the present study, we suggest that these might be fruitful areas for future research.

Finally, we would like to remark on the results regarding Spiritist churches, which in some way are the mirror image of those for Evangelical churches. An extensive scan of the literature finds much less information about this religious tradition, in particular in the context of Brazil where it adapted in ways that made it very different from its original French predecessor [see @daGraca2011espiritismo]. Could it be that Spiritism is seen by potential criminals as spiritually or materially risky, if supernatural retribution is feared? Again, we cannot do more than speculate about this at the moment, but this might be another worthwhile avenue for future investigation.

@@

1. Catholic places of worship do not seem to geographically project more or less of a moral community effect than, say, ice cream shops, pharmacies, or supermarkets: the intensity of crime as a function of distance to Catholic churches is not significantly different that what we observe when we consider distance to any of these types of commercial establishments, at any distance.

2. The intensity of crime with respect to distance to Evangelical churches is more complex than previously known: whereas the intensity of crime is higher at relatively short distance from Evangelical places of worship, at longer distances it tends to decline even below the global intensity of crime and the background level of crime according to Socio-Economic Deprivation.

@@

In summary, the research presented on this paper provides information about the potential of different types of moral communities to reduce crime, and should be of interest to policy makers as they assess whether formal or informal forms of social control can be effective to deter criminal behavior, in order to achieve development goals.

# Acknowledgments

The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `faterize` [@Ross2018fasterize], `ggthemes` [@Arnold2018], `kableExtra` [@Zhu2018], `knitr` [@Xie2015; @Xie2018], `maptools` [@Bivand2019maptools], `spatstat` [@Baddeley2005spatstat], `raster` [@Hijmans2018raster], `rticles` [@Allaire2018rticles], `sf` [@Pebesma2018simple], and `tidyverse` [@Wickham2017].
